EEG Waveform Analysis of P300 ERP with applications to Brain Computer Interfaces

### 2018-09-17 sensors.tex
Confirmed section

%=================================================================
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother 
\articlenumber{x}
\doinum{10.3390/------}
\pubvolume{xx}
\pubyear{2017}
\copyrightyear{2017}
\externaleditor{Academic Editor: name}
\history{Received: date; Accepted: date; Published: date}

@@

Waveform characterization is the method in which analysis has been performed for Event Related Potentials.  Evoked Potentials (EPs) and Event Related Potentials (ERPs) are transient signal elements that may arise as a brain response to an external visual, tactile or auditory stimulus.  Particularly, ERPs are regularly used to assess auditory response in infants.  They are extensively used and studied in Cognitive Neuroscience~\citep{Luck2005}.  ERPs are identified by their components which are recognizable signal shapes assigned to the observed waveform, that can be linked to some cognitive or measurable psychological process.   The P300 is one of the most studied ERP component in BCI because it can be used to implement a Speller application.  Hence, P300 ERPs are a target phenomena to study by automatic waveform recognition methods.

@@

One of the earliest approach to automatically process EEG data is the Peak Picking method.  Although of limited usability, peak picking has been used to determine latency of transient events in EEG \citep{Jaskowski2000,Zhang2011}.  Straightforward in its implementation, it consists in assigning a component to a simple waveform element based on the expected location of its more prominent deflection \citep{Ouyang2017}.  Evoked Potentials (EPs) and Event Related Potentials (ERPs) are transient component that may arise as a brain response to an external visual, tactile or auditory stimulus.  Particularly, ERPs are regularly used to assess auditory response in infants. They are characterized in this manner, where the name of many of the EEG features evoke directly a peak within the waveform, e.g. P300 or P3a P3b or N100.  This leads to a natural procedure to classify them visually by selecting appropriate peaks and matching their positions and amplitudes in an orderly manner.  The letter provides the polarity (Positive or Negative) and the numbering shows the time referencing the stimulus onset, or the ordinal position of each peak (first, second, etc).

@@

All these methods provide a feature $ f $  that can be used as a template.  The notation $()_{1}^{n} $ will be used to describe the concatenation of scalar values to form a multidimensional feature vector $f$. Additionally, thy are based on metrics that are extracted from the shape of the single channel digital EEG signal $x(n)$, with $n$ varying from $1$ to the length $N$ of the EEG segment in sample points.  These features are used to create dictionaries or template databases.  Finally, these templates provide the basis for the pattern matching algorithm and offline classification. These algorithm were implemented on MATLAB 2014a (Mathworks Inc., Natick, MA, USA).

@@

\textit{Pursuit} algorithms refer, in their many variants, as blind source separation \citep{Vincent2010} techniques that assume the EEG signal as a linear combination of different sparse sources extracted from a template's dictionaries.  Matching Pursuit \textit{MP}~\citep{Mallat1993}, the most representative of this algorithms, is a greedy variant that decomposes a signal into a linear combination of waveforms, called atoms, that are well localized in time and frequency \citep{ChandranKS2016}.  Given a signal, this optimization technique, tries to find the indexes of atoms and their weights (contributions) that minimize,

@@

which is the error between the signal and their approximation constructed by the weighted $w_i$ atoms $g_{i}$, and calculating the euclidean norm ${\left\lVert \cdot \right\rVert}_{2}$.  The algorithm goes by first setting the approximating signal $\tilde{x}_{0}$  as the original signal itself,

@@

where $g_{i}$ are all the available scaled, translated and modulated atoms from the dictionary.  The operation $\left\lvert \cdot \right\rvert$ corresponds to the absolute value of the inner product.  This step determines the atom selection process, and their contribution is calculated based on

@@

with $k$ representing the index of the selected atom $g_{k}$.  Finally the contribution of each atom is subtracted from the next approximation~\citep{Cohen2014,Sanei2007, Mallat1993}

\begin{equation}
\tilde{x}_{k+1}(n)=  \tilde{x}_{k}(n) - w_{k} g_{k} (n)
\label{eq:mp5}
\end{equation}

@@

\begin{equation}
f =  {\bigg ( w_{i} \bigg )}_{1}^{D} 
\label{eq:mp6}
\end{equation}

@@

\begin{equation}
f =  {\bigg ( H(\textbf{p})_{u} \bigg )}_{1}^{N - (W + \tau (m - 1)) + 1}
\label{eq:pe6}
\end{equation}

@@

\begin{equation}
f = \bigg (  \frac{\hat{y}_{s}-\hat{y}_{s-1}}{\hat{x}_{s}-\hat{x}_{s-1}}  \bigg )_{2}^{G} 
\label{eq:shcc8}
\end{equation}

@@

\begin{equation}
f = {\bigg ( h(i,j,\theta) \bigg )}_{(0,0,0)}^{(3,3,315)}
\label{eq:sift7}
\end{equation}

>>>>

%=================================================================
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother 
\articlenumber{x}
\doinum{10.3390/------}
\pubvolume{xx}
\pubyear{2018}
\copyrightyear{2018}
\externaleditor{Academic Editor: name}
\history{Received: date; Accepted: date; Published: date}

@@

Waveform characterization is the method in which analysis has been performed for Event Related Potentials.  Evoked Potentials (EPs) and Event Related Potentials (ERPs) are transient signal elements that may arise as a brain response to an external visual, tactile or auditory stimulus.  ERPs are regularly used to assess auditory response in infants.  They are extensively used and studied in Cognitive Neuroscience~\citep{Luck2005}.  ERPs are identified by their components which are recognizable signal shapes assigned to the observed waveform, that can be linked to some cognitive or measurable psychological process.   The P300 is one of the most studied ERP component in BCI because it can be used to implement a Speller application.  Hence, P300 ERPs are a target phenomena to study by automatic waveform recognition methods.

@@

One of the earliest approach to automatically process EEG data is the Peak Picking method.  Although of limited usability, peak picking has been used to determine latency of transient events in EEG \citep{Jaskowski2000,Zhang2011}.  Straightforward in its implementation, it consists in assigning a component to a simple waveform element based on the expected location of its more prominent deflection \citep{Ouyang2017}.  
Of regular use in ERP Research, the name of many of the EEG features reference directly a peak within the component, e.g. P300 or P3a P3b or N100.  This leads to a natural way to classify them visually by selecting appropriate peaks and matching their positions and amplitudes in an orderly manner.  The letter provides the polarity (Positive or Negative) and the numbering shows the time referencing the stimulus onset, or the ordinal position of each peak (first, second, etc).

%Evoked Potentials (EPs) and Event Related Potentials (ERPs) are transient component that may arise as a brain response to an external visual, tactile or auditory stimulus.  Particularly, ERPs are regularly used to assess auditory response in infants. They are characterized in this manner, where the name of many of the EEG features evoke directly a peak within the waveform, e.g. P300 or P3a P3b or N100.  This leads to a natural procedure to classify them visually by selecting appropriate peaks and matching their positions and amplitudes in an orderly manner.  The letter provides the polarity (Positive or Negative) and the numbering shows the time referencing the stimulus onset, or the ordinal position of each peak (first, second, etc).

@@

All these methods provide a feature $ f $  that can be used as a template.  The notation $\{\}_{1}^{n} $ will be used to describe the concatenation of scalar values to form a multidimensional feature vector $f$. These algorithms are based on metrics that are extracted from the shape of the single channel digital EEG signal $x(n)$, with $n$ varying from $1$ to the length $N$ of the EEG segment in sample points.  These features are used to create dictionaries or template databases.  Finally, these templates provide the basis for the pattern matching algorithm and offline classification. These algorithm were implemented on MATLAB 2014a (Mathworks Inc., Natick, MA, USA).

@@

\textit{Pursuit} algorithms refer, in their many variants, as blind source separation \citep{Vincent2010} techniques that assume the EEG signal as a linear combination of different sparse sources extracted from a template's dictionaries.  Matching Pursuit \textit{MP}~\citep{Mallat1993}, the most representative of this algorithms, is a greedy variant that decomposes a signal into a linear combination of waveforms, called atoms, that are well localized in time and frequency \citep{ChandranKS2016}.  Given a signal, this optimization technique, tries to find the indexes of $m$ atoms and their weights (contributions) that minimize,

@@

\noindent which is the error between the signal and their approximation constructed by the weighted $w_i$ atoms $g_{i}$, and calculating the euclidean norm ${\left\lVert \cdot \right\rVert}_{2}$.  The algorithm goes by first setting the approximating signal $\tilde{x}_{0}$  as the original signal itself,

@@

\noindent where $g_{i}$ are all the available scaled, translated and modulated atoms from the dictionary.  The operation $\left\lvert \cdot \right\rvert$ corresponds to the absolute value of the inner product.  This step determines the atom selection process, and their contribution is calculated based on

@@

\noindent with $k$ representing the index of the selected atom $g_{k}$ and $\left\lVert \cdot \right\rVert$ its norm.  Finally the contribution of each atom is subtracted from the next approximation~\citep{Cohen2014,Sanei2007, Mallat1993}

\begin{equation}
\tilde{x}_{k}(n)=  \tilde{x}_{k-1}(n) - w_{k} g_{k} (n)
\label{eq:mp5}
\end{equation}

@@

\begin{equation}
f =  {\bigg \{ w_{i} \bigg \}}_{1}^{D} 
\label{eq:mp6}
\end{equation}

@@

\begin{equation}
f =  {\bigg \{ H(\textbf{p})_{u} \bigg \}}_{1}^{N - (W + \tau (m - 1)) + 1}
\label{eq:pe6}
\end{equation}

@@

\begin{equation}
f = \bigg \{  \frac{\hat{y}_{s}-\hat{y}_{s-1}}{\hat{x}_{s}-\hat{x}_{s-1}}  \bigg \}_{2}^{G} 
\label{eq:shcc8}
\end{equation}

@@

\begin{equation}
f = {\bigg \{ h(i,j,\theta) \bigg \}}_{(0,0,0)}^{(3,3,315)}
\label{eq:sift7}
\end{equation}
